---
title: "Gene expression analysis of keloid tissue and cultured fibroblasts"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

Source codes: https://github.com/markziemann/rnaseq-keloid

## Background

Here we are looking at the molecular hallmarks of Keloids in tissue samples and fibroblasts.

Reads underwent quality trimming using Skewer version 0.2.2 (Jiang et al, 2014).
I mapped the reads to the Gencode transcriptome version 35 using Kallisto version 0.46.1 (Bray et al, 2016).
Expression counts were loaded into R and then DE analysis was performed with DESeq2 (Love et al, 2014).
Enrichment analysis was performed using Reactome and MSigDB genesets (Jassal et al, 2020; Liberzon et al, 2015) with the Mitch package (Kaspi & Ziemann 2020).

```{r,pkg}

suppressPackageStartupMessages({
    library("reshape2")
    library("gplots")
    library("DESeq2")
    library("mitch")
})

```

Here are some functions.

```{r,defunc}

run_de <- function(ss,xx){

y <- round(xx)

# MDS
mds <- cmdscale(dist(t(y)))
XMAX=max(mds[,1])*1.1
XMIN=min(mds[,1])*1.1
plot( mds , xlab="Coordinate 1", ylab="Coordinate 2",
  type = "n" , xlim=c(XMIN,XMAX),main="MDS plot",bty="n")
text(mds, labels=colnames(y) )

# DE
dds <- DESeqDataSetFromMatrix(countData=y, colData = ss, design = ~ trt)
dds <- DESeq(dds)
de <- DESeq2::results(dds)
de <- de[order(de$pvalue),]
up <- rownames(subset(de, log2FoldChange>0 & padj<0.05 ))
dn <- rownames(subset(de, log2FoldChange<0 & padj<0.05 ))
str(up)
str(dn)

# MA plot
sig <-subset(de, padj < 0.05 )
GENESUP <- length(up)
GENESDN <- length(dn)
SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down")
ns <-subset(de, padj > 0.05 )
plot(log2(de$baseMean),de$log2FoldChange,
     xlab="log2 basemean", ylab="log2 foldchange",
     pch=19, cex=0.5, col="dark gray",
     main="smear plot")
points(log2(sig$baseMean),sig$log2FoldChange,
       pch=19, cex=0.5, col="red")
mtext(SUBHEADER)

# heatmap
yn <- y/colSums(y)*1000000
yf <- yn[which(rownames(yn) %in% rownames(de)[1:50]),]
mycols <- gsub("0","yellow",ss$trt)
mycols <- gsub("1","orange",mycols)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(yf), col=colfunc(25),scale="row",
    ColSideColors =mycols ,trace="none",
    margin = c(10,10), cexRow=0.6, cexCol=0.8 , main="Top 50 genes by p-val")
mtext("yellow=ctrl, orange=trt")

return(de)
}

```


## Load data

Here we load the data in from the aligner.

```{r,import}

tmp <- read.table("3col.tsv.gz")
x <- as.data.frame(acast(tmp, V2~V1, value.var="V3"))
x$geneID <- sapply(strsplit(rownames(x),"\\|"),"[[",2)
x$gene_name <- sapply(strsplit(rownames(x),"\\|"),"[[",6)
x$gene <- paste(x$geneID,x$gene_name)
x$geneID = x$gene_name = NULL
xx <- aggregate(. ~ gene, x, sum)
rownames(xx) <- xx$gene
xx$gene = NULL
head(xx)

```

```{r,numreads,fig.width=7,fig.height=7}

par(mar=c(5,10,5,3))
barplot(colSums(xx),horiz=TRUE,las=2,main="number of reads per sample",cex.names=0.5)


```

## Sample sheet

The data were divided into two study arms: tissues and cells.
Here are the contrasts for the tissues where group 1 is case and group 2 is control.

        Group1          Group2          Check

1       KP (35,45,46)   N (6,9,10)      missing KP45

2       KM (35,45,46)   N (6,9,10)      OK

3       KT (35,45,46)   N (6,9,10)      OK

4       KN (35,45,46)   N (6,9,10)      OK

5       D (1,3,4)       N (6,9,10)      OK

K=keloid

KP=keloid-peripheral

KM=keloid-middle

KT=keloid-transitional

KN=keloid-normal 

N=normal skin control

I noticed that for group KP, there was only data 35 and 46 present but no 45.
This sample will be included in subsequent analysis.

Here are the contrasts for the cells:

6       KDUP-Peri (35,45,46)    NBF (6,9,10)            OK

7       KDUP-Mid (35,45,46)     NBF (6,9,10)            OK

8       KDUP-T (35,45,46)       NBF (6,9,10)            OK

9       KDUP-N (35,45,46)       NBF (6,9,10)            OK

10      DFSP (CL-2, 3,4)        NBF (6,9,10)            OK

11      FS-CL (1, 2)            NBF (6,9,10)            OK

12      PT-K (Peri, Mid, T)     NBF (6,9,10)            OK

13      PT-K (Peri, Mid, T)     KDUP (Peri, Mid, T, N)  OK

14      PT-K (Peri, Mid, T)     DFSP (CL-2, 3,4)        OK

15      PT-K (Peri, Mid, T)     FS-CL (1, 2)            OK

KDUP-Peri=Keloid Peripheral fibroblasts

KDUP-T=Keloid Transitional fibroblasts

KDUP-Mid= Keloid Middle fibroblasts

KDUP-N=Keloid Normal skin fibroblasts

NF=Normal fibroblast

DFSP=Dermatofibrosarcoma protuberans


```{r,samplesheet}

ss <- data.frame(colnames(xx))
rownames(ss) <- ss[,1]

# specify groups
ss$n <- as.integer(grepl("^N-",ss[,1]))
ss$kp <- as.integer(grepl("^KP",ss[,1]))
ss$km <- as.integer(grepl("^KM",ss[,1]))
ss$kt <- as.integer(grepl("^KT",ss[,1]))
ss$kn <- as.integer(grepl("^KN",ss[,1]))
ss$d <- as.integer(grepl("^D-",ss[,1]))

ss$nbf <- as.integer(grepl("^NBF",ss[,1]))
ss$kdup_peri <- as.integer(grepl("^KDUP-Peri",ss[,1]))
ss$kdup_mid <- as.integer(grepl("^KDUP-Mid",ss[,1]))
ss$kdup_t <- as.integer(grepl("^KDUP-T",ss[,1]))
ss$kdup_n <- as.integer(grepl("^KDUP-N",ss[,1]))
ss$dfsp <- as.integer(grepl("^DFSP",ss[,1]))
ss$fs <- as.integer(grepl("^FS",ss[,1]))
ss$ptk <- as.integer(grepl("^PT-K",ss[,1]))

ss[,1]=NULL

```

## MDS

MDS is just like PCA.
The more similar (correlated) the data sets are the closer they will appear on the 
scatterplot.

```{r,mds,fig.height=8,fig.width=8}

mds <- cmdscale(dist(t(xx)))
XMAX=max(mds[,1])*1.1
XMIN=min(mds[,1])*1.1
plot( mds , xlab="Coordinate 1", ylab="Coordinate 2", 
  type = "n" , xlim=c(XMIN,XMAX),main="MDS plot",bty="n")
text(cmdscale(dist(t(xx))), labels=colnames(xx) )
 
```

We can also visualise the similarities and differences between samples with a heatmap of 
Pearson correlation values.

```{r,cor,fig.height=8,fig.width=8}

colfunc <- colorRampPalette(c("white", "yellow","orange","red", "darkred"))
# Dataset level correlation
heatmap.2(cor(xx),trace="none",margin=c(10,10),scale="none",col=colfunc,
cexRow=0.8,cexCol=0.8)

```

## Split data into contrasts for separate analysis

```{r,split}

# tissues
ss1 <- subset(ss,kp==1|n==1)
ss1$trt <- ss1$kp
xx1 <- xx[,which(colnames(xx) %in% rownames(ss1))] 

ss2 <- subset(ss,km==1|n==1)
ss2$trt <- ss2$km
xx2 <- xx[,which(colnames(xx) %in% rownames(ss2))]

ss3 <- subset(ss,kt==1|n==1)
ss3$trt <- ss3$kt
xx3 <- xx[,which(colnames(xx) %in% rownames(ss3))]

ss4 <- subset(ss,kn==1|n==1)
ss4$trt <- ss4$kn
xx4 <- xx[,which(colnames(xx) %in% rownames(ss4))]

ss5 <- subset(ss,d==1|n==1)
ss5$trt <- ss5$d
xx5 <- xx[,which(colnames(xx) %in% rownames(ss5))]

# cells
ss6 <- subset(ss,kdup_peri==1|nbf==1)
ss6$trt <- ss6$kdup_peri
xx6 <- xx[,which(colnames(xx) %in% rownames(ss6))]

ss7 <- subset(ss,kdup_mid==1|nbf==1)
ss7$trt <- ss7$kdup_mid
xx7 <- xx[,which(colnames(xx) %in% rownames(ss7))]

ss8 <- subset(ss,kdup_t==1|nbf==1)
ss8$trt <- ss8$kdup_t
xx8 <- xx[,which(colnames(xx) %in% rownames(ss8))]

ss9 <- subset(ss,kdup_n==1|nbf==1)
ss9$trt <- ss9$kdup_n
xx9 <- xx[,which(colnames(xx) %in% rownames(ss9))]

ss10 <- subset(ss,dfsp==1|nbf==1)
ss10$trt <- ss10$dfsp  
xx10 <- xx[,which(colnames(xx) %in% rownames(ss10))]

ss11 <- subset(ss,fs==1|nbf==1)
ss11$trt <- ss11$fs
xx11 <- xx[,which(colnames(xx) %in% rownames(ss11))]

ss12 <- subset(ss,ptk==1|nbf==1)
ss12$trt <- ss12$ptk
xx12 <- xx[,which(colnames(xx) %in% rownames(ss12))]

# check whether this one should be "paired"
ss13 <- subset(ss,ptk==1|kdup_peri==1|kdup_mid==1|kdup_t==1|kdup_n==1)
ss13$trt <- ss13$ptk
xx13 <- xx[,which(colnames(xx) %in% rownames(ss13))]

ss14 <- subset(ss,ptk==1|dfsp==1)
ss14$trt <- ss14$ptk
xx14 <- xx[,which(colnames(xx) %in% rownames(ss14))]

ss15 <- subset(ss,ptk==1|fs==1)
ss15$trt <- ss15$ptk
xx15 <- xx[,which(colnames(xx) %in% rownames(ss15))]

```

## DE

Here, were using DESeq2 to perform differential expression analysis for the specified 
contrasts.
The run_de function does the analysis and generate the charts.
Here we actually run the analysis.

```{r,de1,fig.height=8,fig.width=8}

ss1
de1 <- run_de(ss1,xx1)
head(de1)
write.table(de1,file="de1.tsv",quote=FALSE,sep="\t")

ss2
de2 <- run_de(ss2,xx2)
head(de2)
write.table(de2,file="de2.tsv",quote=FALSE,sep="\t")

ss3
de3 <- run_de(ss3,xx3)
head(de3)
write.table(de3,file="de3.tsv",quote=FALSE,sep="\t")

ss4
de4 <- run_de(ss4,xx4)
head(de4)
write.table(de4,file="de4.tsv",quote=FALSE,sep="\t")

ss5
de5 <- run_de(ss5,xx5)
head(de5)
write.table(de5,file="de5.tsv",quote=FALSE,sep="\t")

ss6
de6 <- run_de(ss6,xx6)
head(de6)
write.table(de6,file="de6.tsv",quote=FALSE,sep="\t")

ss7
de7 <- run_de(ss7,xx7)
head(de7)
write.table(de7,file="de7.tsv",quote=FALSE,sep="\t")

ss8
de8 <- run_de(ss8,xx8)
head(de8)
write.table(de8,file="de8.tsv",quote=FALSE,sep="\t")

ss9
de9 <- run_de(ss9,xx9)
head(de9)
write.table(de9,file="de9.tsv",quote=FALSE,sep="\t")

ss10
de10 <- run_de(ss10,xx10)
head(de10)
write.table(de10,file="de10.tsv",quote=FALSE,sep="\t")

ss11
de11 <- run_de(ss11,xx11)
head(de1)
write.table(de11,file="de11.tsv",quote=FALSE,sep="\t")

ss12
de12 <- run_de(ss12,xx12)
head(de12)
write.table(de12,file="de12.tsv",quote=FALSE,sep="\t")

ss13
de13 <- run_de(ss13,xx13)
head(de13)
write.table(de13,file="de13.tsv",quote=FALSE,sep="\t")

ss14
de14 <- run_de(ss14,xx14)
head(de14)
write.table(de14,file="de14.tsv",quote=FALSE,sep="\t")

ss15
de15 <- run_de(ss15,xx15)
head(de15)
write.table(de15,file="de15.tsv",quote=FALSE,sep="\t")

```

## Pathway analysis

Will do MSigDB and Reactome enrichment in mitch.
Reactome gene sets were downloaded 19 Oct 2020.
I used MSigDB version 7.2.


```{r,mitch,fig.height=8,fig.width=8}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
reactome <- gmt_import("ReactomePathways.gmt")
msigdb <- gmt_import("msigdb.v7.2.symbols.gmt")

gt <- as.data.frame(rownames(xx))

gt$gn <- sapply(strsplit(gt[,1]," "),"[[",2)

mitch_barplot <- function(res){
  sig <- head(subset(res$enrichment_result,p.adjustANOVA<0.05),30)
  sig <- sig[order(sig$s.dist),]
  par(mar=c(3,25,1,1)); barplot(sig$s.dist,horiz=TRUE,las=2,cex.names = 0.6,cex.axis = 0.6,
    names.arg=sig$set,main="Enrichment score") ;grid()
}

m1 <- mitch_import(x=as.data.frame(de1),DEtype="deseq2",geneTable=gt)
res1 <- mitch_calc(x=m1,genesets=reactome,priority="effect")
head(res1$enrichment_result,20)
mitch_barplot(res1)
unlink("de1_mitch_reactome.html")
mitch_report(res1,outfile="de1_mitch_reactome.html")
res1m <- mitch_calc(x=m1,genesets=msigdb,priority="effect",cores=12)
head(res1m$enrichment_result,20)
mitch_barplot(res1m)
unlink("de1_mitch_msigdb.html")
mitch_report(res1m,outfile="de1_mitch_msigdb.html")

m2 <- mitch_import(x=as.data.frame(de2),DEtype="deseq2",geneTable=gt)
res2 <- mitch_calc(x=m2,genesets=reactome,priority="effect")
head(res2$enrichment_result,20)
mitch_barplot(res2)
unlink("de2_mitch_reactome.html")
mitch_report(res2,outfile="de2_mitch_reactome.html")
res2m <- mitch_calc(x=m2,genesets=msigdb,priority="effect",cores=12)
head(res2m$enrichment_result,20)
mitch_barplot(res2m)
unlink("de2_mitch_msigdb.html")
mitch_report(res2m,outfile="de2_mitch_msigdb.html")

m3 <- mitch_import(x=as.data.frame(de3),DEtype="deseq2",geneTable=gt)
res3 <- mitch_calc(x=m3,genesets=reactome,priority="effect")
head(res3$enrichment_result,20)
mitch_barplot(res3)
unlink("de3_mitch_reactome.html")
mitch_report(res3,outfile="de3_mitch_reactome.html")
res3m <- mitch_calc(x=m3,genesets=msigdb,priority="effect",cores=12)
head(res3m$enrichment_result,20)
mitch_barplot(res3m)
unlink("de3_mitch_msigdb.html")
mitch_report(res3m,outfile="de3_mitch_msigdb.html")

m4 <- mitch_import(x=as.data.frame(de4),DEtype="deseq2",geneTable=gt)
res4 <- mitch_calc(x=m4,genesets=reactome,priority="effect")
head(res4$enrichment_result,20)
mitch_barplot(res4)
unlink("de4_mitch_reactome.html")
mitch_report(res4,outfile="de4_mitch_reactome.html")
res4m <- mitch_calc(x=m4,genesets=msigdb,priority="effect",cores=12)
head(res4m$enrichment_result,20)
mitch_barplot(res4m)
unlink("de4_mitch_msigdb.html")
mitch_report(res4m,outfile="de4_mitch_msigdb.html")

m5 <- mitch_import(x=as.data.frame(de5),DEtype="deseq2",geneTable=gt)
res5 <- mitch_calc(x=m5,genesets=reactome,priority="effect")
head(res5$enrichment_result,20)
mitch_barplot(res5)
unlink("de5_mitch_reactome.html")
mitch_report(res5,outfile="de5_mitch_reactome.html")
res5m <- mitch_calc(x=m5,genesets=msigdb,priority="effect",cores=12)
head(res5m$enrichment_result,20)
mitch_barplot(res5m)
unlink("de5_mitch_msigdb.html")
mitch_report(res5m,outfile="de5_mitch_msigdb.html")

m6 <- mitch_import(x=as.data.frame(de6),DEtype="deseq2",geneTable=gt)
res6 <- mitch_calc(x=m6,genesets=reactome,priority="effect")
head(res6$enrichment_result,20)
mitch_barplot(res6)
unlink("de6_mitch_reactome.html")
mitch_report(res6,outfile="de6_mitch_reactome.html")
res6m <- mitch_calc(x=m6,genesets=msigdb,priority="effect",cores=12)
head(res6m$enrichment_result,20)
mitch_barplot(res6m)
unlink("de6_mitch_msigdb.html")
mitch_report(res6m,outfile="de6_mitch_msigdb.html")

m7 <- mitch_import(x=as.data.frame(de7),DEtype="deseq2",geneTable=gt)
res7 <- mitch_calc(x=m7,genesets=reactome,priority="effect")
head(res7$enrichment_result,20)
mitch_barplot(res7)
unlink("de7_mitch_reactome.html")
mitch_report(res7,outfile="de7_mitch_reactome.html")
res7m <- mitch_calc(x=m7,genesets=msigdb,priority="effect",cores=12)
head(res7m$enrichment_result,20)
mitch_barplot(res7m)
unlink("de7_mitch_msigdb.html")
mitch_report(res7m,outfile="de7_mitch_msigdb.html")

m8 <- mitch_import(x=as.data.frame(de8),DEtype="deseq2",geneTable=gt)
res8 <- mitch_calc(x=m8,genesets=reactome,priority="effect")
head(res8$enrichment_result,20)
mitch_barplot(res8)
unlink("de8_mitch_reactome.html")
mitch_report(res8,outfile="de8_mitch_reactome.html")
res8m <- mitch_calc(x=m8,genesets=msigdb,priority="effect",cores=12)
head(res8m$enrichment_result,20)
mitch_barplot(res8m)
unlink("de8_mitch_msigdb.html")
mitch_report(res8m,outfile="de8_mitch_msigdb.html")

m9 <- mitch_import(x=as.data.frame(de9),DEtype="deseq2",geneTable=gt)
res9 <- mitch_calc(x=m9,genesets=reactome,priority="effect")
head(res9$enrichment_result,20)
mitch_barplot(res9)
unlink("de9_mitch_reactome.html")
mitch_report(res9,outfile="de9_mitch_reactome.html")
res9m <- mitch_calc(x=m9,genesets=msigdb,priority="effect",cores=12)
head(res9m$enrichment_result,20)
mitch_barplot(res9m)
unlink("de9_mitch_msigdb.html")
mitch_report(res9m,outfile="de9_mitch_msigdb.html")

m10 <- mitch_import(x=as.data.frame(de10),DEtype="deseq2",geneTable=gt)
res10 <- mitch_calc(x=m10,genesets=reactome,priority="effect")
head(res10$enrichment_result,20)
mitch_barplot(res10)
unlink("de10_mitch_reactome.html")
mitch_report(res10,outfile="de10_mitch_reactome.html")
res10m <- mitch_calc(x=m10,genesets=msigdb,priority="effect",cores=12)
head(res10m$enrichment_result,20)
mitch_barplot(res10m)
unlink("de10_mitch_msigdb.html")
mitch_report(res10m,outfile="de10_mitch_msigdb.html")

m11 <- mitch_import(x=as.data.frame(de11),DEtype="deseq2",geneTable=gt)
res11 <- mitch_calc(x=m11,genesets=reactome,priority="effect")
head(res11$enrichment_result,20)
mitch_barplot(res11)
unlink("de11_mitch_reactome.html")
mitch_report(res11,outfile="de11_mitch_reactome.html")
res11m <- mitch_calc(x=m11,genesets=msigdb,priority="effect",cores=12)
head(res11m$enrichment_result,20)
mitch_barplot(res11m)
unlink("de11_mitch_msigdb.html")
mitch_report(res11m,outfile="de11_mitch_msigdb.html")

m12 <- mitch_import(x=as.data.frame(de12),DEtype="deseq2",geneTable=gt)
res12 <- mitch_calc(x=m12,genesets=reactome,priority="effect")
head(res12$enrichment_result,20)
mitch_barplot(res12)
unlink("de12_mitch_reactome.html")
mitch_report(res12,outfile="de12_mitch_reactome.html")
res12m <- mitch_calc(x=m12,genesets=msigdb,priority="effect",cores=12)
head(res12m$enrichment_result,20)
mitch_barplot(res12m)
unlink("de12_mitch_msigdb.html")
mitch_report(res12m,outfile="de12_mitch_msigdb.html")

m13 <- mitch_import(x=as.data.frame(de13),DEtype="deseq2",geneTable=gt)
res13 <- mitch_calc(x=m13,genesets=reactome,priority="effect")
head(res13$enrichment_result,20)
mitch_barplot(res13)
unlink("de13_mitch_reactome.html")
mitch_report(res13,outfile="de13_mitch_reactome.html")
res13m <- mitch_calc(x=m13,genesets=msigdb,priority="effect",cores=12)
head(res13m$enrichment_result,20)
mitch_barplot(res13m)
unlink("de13_mitch_msigdb.html")
mitch_report(res13m,outfile="de13_mitch_msigdb.html")

m14 <- mitch_import(x=as.data.frame(de14),DEtype="deseq2",geneTable=gt)
res14 <- mitch_calc(x=m14,genesets=reactome,priority="effect")
head(res14$enrichment_result,20)
mitch_barplot(res14)
unlink("de14_mitch_reactome.html")
mitch_report(res14,outfile="de14_mitch_reactome.html")
res14m <- mitch_calc(x=m14,genesets=msigdb,priority="effect",cores=12)
head(res14m$enrichment_result,20)
mitch_barplot(res14m)
unlink("de14_mitch_msigdb.html")
mitch_report(res14m,outfile="de14_mitch_msigdb.html")

m15 <- mitch_import(x=as.data.frame(de15),DEtype="deseq2",geneTable=gt)
res15 <- mitch_calc(x=m15,genesets=reactome,priority="effect")
head(res15$enrichment_result,20)
mitch_barplot(res15)
unlink("de15_mitch_reactome.html")
mitch_report(res15,outfile="de15_mitch_reactome.html")
res15m <- mitch_calc(x=m15,genesets=msigdb,priority="effect",cores=12)
head(res15m$enrichment_result,20)
mitch_barplot(res15m)
unlink("de15_mitch_msigdb.html")
mitch_report(res15m,outfile="de15_mitch_msigdb.html")

save.image("stats.Rdata")

```

## Session information

So you know what version of R and packages was used.

```{r,session}
sessionInfo()
```

## References

Bray NL, Pimentel H, Melsted P, Pachter L. Near-optimal probabilistic RNA-seq quantification [published correction appears in Nat Biotechnol. 2016 Aug 9;34(8):888]. Nat Biotechnol. 2016;34(5):525-527. doi:10.1038/nbt.3519

Jiang H, Lei R, Ding SW, Zhu S. Skewer: a fast and accurate adapter trimmer for next-generation sequencing paired-end reads. BMC Bioinformatics. 2014;15:182. Published 2014 Jun 12. doi:10.1186/1471-2105-15-182

Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol. 2014;15(12):550. doi:10.1186/s13059-014-0550-8

Jassal B, Matthews L, Viteri G, Gong C, Lorente P, Fabregat A, Sidiropoulos K, Cook J, Gillespie M, Haw R, Loney F, May B, Milacic M, Rothfels K, Sevilla C, Shamovsky V, Shorser S, Varusai T, Weiser J, Wu G, Stein L, Hermjakob H, D'Eustachio P. The reactome pathway knowledgebase. Nucleic Acids Res. 2020 Jan 8;48(D1):D498-D503. doi: 10.1093/nar/gkz1031. PMID: 31691815; PMCID: PMC7145712.

Liberzon A, Birger C, Thorvaldsdóttir H, Ghandi M, Mesirov JP, Tamayo P. The Molecular Signatures Database (MSigDB) hallmark gene set collection. Cell Syst. 2015 Dec 23;1(6):417-425. doi: 10.1016/j.cels.2015.12.004. PMID: 26771021; PMCID: PMC4707969.

Kaspi A, Ziemann M. mitch: multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics. 2020;21(1):447. Published 2020 Jun 29. doi:10.1186/s12864-020-06856-9

